Parsing Postfix log files

### 2009-03-26 doc/logparser-complications.tex
Added a \action{} command to have consistent formatting of action names. Changes \parsernames{} to have an ' before the s.

Tracking re-injected mail requires \parsername{} to do the following in the
COMMIT action:

@@

This check is performed during the COMMIT action.

@@

    \item Was the second result processed by the \texttt{CLONE} action?
        The first two \daemon{smtpd} log lines will be a connection log
        line and a mail acceptance log line.

@@

    \item If there are third and subsequent results, were those log lines
        processed by the \texttt{SAVE\_DATA} action?  If there are any log
        lines after the first two they should be informational only.

@@

This check is performed in the DISCONNECT action; it requires support in
the CLONE action where the list of accepted mails is maintained.

@@

These checks are made during the DISCONNECT action.

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes there is a timeout or the
connection is lost during the DATA phase;\footnote{For the sake of brevity
\textit{timeout\/} will be used throughout this section, but everything
applies equally to lost connections.} when this occurs Postfix will discard
the mail and \parsername{} needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded and saved with the connection instead.  To deal properly with
timeouts the TIMEOUT action does the following:

@@

    \item If no mails have been accepted yet nothing needs to be done; the
        TIMEOUT action ends.

@@

        XXX THE REMAINDER OF THIS DOES NOT MAKE SENSE\@; CHECK AND CORRECT
        IT\@.  To distinguish between the two cases the action compares the
        timestamp of the last accepted mail against the timestamp of the
        last log line from \daemon{smtpd} for that connection; this depends
        on the CLONE action keeping a list of all mails accepted for each
        connection.  If the mail acceptance timestamp is later than the
        timeout timestamp, the timeout applies to the just-accepted mail,
        which will be removed from the state tables, without entering it in
        the database.  If the \daemon{smtpd} timestamp is later there was a
        rejection between the accepted mail and the timeout: the action
        assumes that the timeout applies to the rejected delivery attempt
        and finishes.  This assumption is not necessarily correct, because
        Postfix may have accepted an earlier recipient and rejected a later
        one, in which case the timeout applies to the accepted mail, which
        should be discarded; this has not been a problem in practice.
        Processing timeouts is further complicated by the presence of out
        of order \daemon{cleanup} log lines: see \sectionref{discarding
        cleanup log lines} for details.

@@

This complication is dealt with by the TIMEOUT action, with help from the
CLONE action.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} log line for a mail being accepted is logged after the
timeout log line, by which time \parsername{} has discarded the mail;
parsing the \daemon{cleanup} log line causes the CLEANUP\_PROCESSING action
to create a new state table entry, to help deal with re-injected mails
(\sectionref{Re-injected mails}).  This is incorrect because the log line
actually belongs to the mail that has just been discarded; if the queueid
is reused there will be a queueid clash, otherwise the new mail will just
remain in the state tables.

During the TIMEOUT action, if the \daemon{cleanup} line is still pending,
the action updates the \texttt{timeout\_queueids} state table, adding the
queueid and the timestamp from the log line.  When the next
\daemon{cleanup} line is parsed for that queueid, the list will be checked
and if the log line meets the following requirements it will be deemed part
of the mail where the timeout occurred and discarded.

@@

The \daemon{cleanup} log line must meet the criteria above for it to be
discarded because some, but not all, connections where a timeout occurs
will have an associated \daemon{cleanup} log line; if the
CLEANUP\_PROCESSING action blindly discarded the next \daemon{cleanup} log
line after a timeout it would sometimes be mistaken.  When the next
\daemon{pickup} log line containing that queueid is processed the queueid
will be removed from the cache of timeout queueids, regardless of whether
it meets the criteria above.

This complication is handled by the CLEANUP\_PROCESSING and TIMEOUT
actions.

@@

When mail is submitted locally, \daemon{pickup} accepts the new mail and
generates a log line.  Occasionally this log line will occur later in the
log file than the \daemon{cleanup} log line, so the PICKUP action will find
that a state table entry already exists for that queueid.  Normally when
this happens a warning is generated by the PICKUP action, but if the
following conditions are met it is assumed that the log lines were out of
order:

@@

This complication is dealt with by the PICKUP action.

@@

These occurrences are handled by the actions POSTFIX\_RELOAD (1),
SMTPD\_DIED (2 \& 3), and SMTPD\_WATCHDOG (4).

@@

The COMMIT action examines the list of programs that have produced log lines
for each mail, comparing the list against a table of known-good program
combinations.  If the mail's combination is found in the table the mail can be
entered in the database; if the combination is not found entry must be
postponed and the mail flagged for later entry.  The MAIL\_DELIVERED action
checks for that flag; if the additional log lines have caused the mail to reach
a valid combination then entry in the database will proceed, otherwise it must
be postponed once more.

@@

This check applies to accepted mails only, not to rejected mails.  This
check is performed during the COMMIT action, with support from the
MAIL\_DELIVERED action.

@@

The DELETE action is responsible for handling this complication.

@@

This is yet another complication that only occurs during periods of
extremely high load, when process scheduling and even hard disk access
times cause strange behaviour.  In this complication, bounce notification
mails are created, delivered, and deleted from the queue, \textit{before\/}
the log line from \daemon{bounce} that explains their source.  To deal with
this the COMMIT action maintains a cache of recently committed bounce
notification mails, which the BOUNCE action subsequently checks if the
bounce mail is not found in the state tables.  If the queueid exists in the
cache, and its start time is less than ten seconds before the timestamp of
the bounce log line, it is assumed that the bounce notification mail has
already been processed and the BOUNCE action does not create one.  If the
queueid exists in the cache it is removed, because it has either just been
used or the problem did not occur for that mail.  Whether the BOUNCE action
creates a mail or finds an existing mail in the state tables, it flags the
mail as having been seen by the BOUNCE action; if this flag is present the
COMMIT action will not add the mail to the cache of recent bounce
notification mails.  This is not required to correctly deal with the
complication, but is an optimisation to reduce the parser's memory usage.
On the occasions the author has observed this complication occurring there
have been a huge number of bounce notification mails generated: if every
bounce notification mail was cached it would dramatically increase the
memory requirements of the parser.  The cache of bounce notification mails
will be pruned whenever the parser's state is saved, though if the size of
the cache ever becomes a problem it could be pruned periodically to keep it
in check.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally
mails that are in the process of being delivered will be deleted.  This
results in the log lines from the delivery agent (\daemon{local},
\daemon{virtual} or \daemon{smtp}) appearing in the log file
\textit{after\/} the mail has been removed from the state tables and saved
in the database.  The DELETE action adds deleted mails to a cache, which is
checked by the SAVE\_DATA action, and the current log line discarded if the
following conditions are met:

>>>>

Tracking re-injected mail requires \parsername{} to do the following in the
\action{COMMIT} action:

@@

This check is performed during the \action{COMMIT} action.

@@

    \item Was the second result processed by the \action{CLONE} action?
        The first two \daemon{smtpd} log lines will be a connection log
        line and a mail acceptance log line.

@@

    \item If there are third and subsequent results, were those log lines
        processed by the \action{SAVE\_DATA} action?  If there are any log
        lines after the first two they should be informational only.

@@

This check is performed in the \action{DISCONNECT} action; it requires
support in the \action{CLONE} action where the list of accepted mails is
maintained.

@@

These checks are made during the \action{DISCONNECT} action.

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes there is a timeout or the
connection is lost during the DATA phase;\footnote{For the sake of brevity
\textit{timeout\/} will be used throughout this section, but everything
applies equally to lost connections.} when this occurs Postfix will discard
the mail and \parsername{} needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded and saved with the connection instead.  To deal properly with
timeouts the \action{TIMEOUT} action does the following:

@@

    \item If no mails have been accepted yet nothing needs to be done; the
        \action{TIMEOUT} action ends.

@@

        XXX THE REMAINDER OF THIS DOES NOT MAKE SENSE\@; CHECK AND CORRECT
        IT\@.  To distinguish between the two cases the action compares the
        timestamp of the last accepted mail against the timestamp of the
        last log line from \daemon{smtpd} for that connection; this depends
        on the \action{CLONE} action keeping a list of all mails accepted
        for each connection.  If the mail acceptance timestamp is later
        than the timeout timestamp, the timeout applies to the
        just-accepted mail, which will be removed from the state tables,
        without entering it in the database.  If the \daemon{smtpd}
        timestamp is later there was a rejection between the accepted mail
        and the timeout: the action assumes that the timeout applies to the
        rejected delivery attempt and finishes.  This assumption is not
        necessarily correct, because Postfix may have accepted an earlier
        recipient and rejected a later one, in which case the timeout
        applies to the accepted mail, which should be discarded; this has
        not been a problem in practice.  Processing timeouts is further
        complicated by the presence of out of order \daemon{cleanup} log
        lines: see \sectionref{discarding cleanup log lines} for details.

@@

This complication is dealt with by the \action{TIMEOUT} action, with help
from the \action{CLONE} action.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} log line for a mail being accepted is logged after the
timeout log line, by which time \parsername{} has discarded the mail;
parsing the \daemon{cleanup} log line causes the
\action{CLEANUP\_PROCESSING} action to create a new state table entry, to
help deal with re-injected mails (\sectionref{Re-injected mails}).  This is
incorrect because the log line actually belongs to the mail that has just
been discarded; if the queueid is reused there will be a queueid clash,
otherwise the new mail will just remain in the state tables.

During the \action{TIMEOUT} action, if the \daemon{cleanup} line is still
pending, the action updates the \texttt{timeout\_queueids} state table,
adding the queueid and the timestamp from the log line.  When the next
\daemon{cleanup} line is parsed for that queueid, the list will be checked
and if the log line meets the following requirements it will be deemed part
of the mail where the timeout occurred and discarded.

@@

The \daemon{cleanup} log line must meet the criteria above for it to be
discarded because some, but not all, connections where a timeout occurs
will have an associated \daemon{cleanup} log line; if the
\action{CLEANUP\_PROCESSING} action blindly discarded the next
\daemon{cleanup} log line after a timeout it would sometimes be mistaken.
When the next \daemon{pickup} log line containing that queueid is processed
the queueid will be removed from the cache of timeout queueids, regardless
of whether it meets the criteria above.

This complication is handled by the \action{CLEANUP\_PROCESSING} and
\action{TIMEOUT} actions.

@@

When mail is submitted locally, \daemon{pickup} accepts the new mail and
generates a log line.  Occasionally this log line will occur later in the
log file than the \daemon{cleanup} log line, so the \action{PICKUP} action
will find that a state table entry already exists for that queueid.
Normally when this happens a warning is generated by the \action{PICKUP}
action, but if the following conditions are met it is assumed that the log
lines were out of order:

@@

This complication is dealt with by the \action{PICKUP} action.

@@

This complication is handled by several actions: \action{POSTFIX\_RELOAD} (1),
\action{SMTPD\_DIED} (2 \& 3), and \action{SMTPD\_WATCHDOG} (4).

@@

The \action{COMMIT} action examines the list of programs that have produced
log lines for each mail, comparing the list against a table of known-good
program combinations.  If the mail's combination is found in the table the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The
\action{MAIL\_DELIVERED} action checks for that flag; if the additional log
lines have caused the mail to reach a valid combination then entry in the
database will proceed, otherwise it must be postponed once more.

@@

This check applies to accepted mails only, not to rejected mails.  This
check is performed during the \action{COMMIT} action, with support from the
\action{MAIL\_DELIVERED} action.

@@

The \action{DELETE} action is responsible for handling this complication.

@@

This is yet another complication that only occurs during periods of
extremely high load, when process scheduling and even hard disk access
times cause strange behaviour.  In this complication, bounce notification
mails are created, delivered, and deleted from the queue, \textit{before\/}
the log line from \daemon{bounce} that explains their source.  To deal with
this the \action{COMMIT} action maintains a cache of recently committed
bounce notification mails, which the \action{BOUNCE\_CREATED} action
subsequently checks if the bounce mail is not found in the state tables.
If the queueid exists in the cache, and its start time is less than ten
seconds before the timestamp of the bounce log line, it is assumed that the
bounce notification mail has already been processed and the
\action{BOUNCE\_CREATED} action does not create one.  If the queueid exists
in the cache it is removed, because it has either just been used or the
problem did not occur for that mail.  Whether the \action{BOUNCE\_CREATED}
action creates a mail or finds an existing mail in the state tables, it
flags the mail as having been seen by the \action{BOUNCE\_CREATED} action;
if this flag is present the \action{COMMIT} action will not add the mail to
the cache of recent bounce notification mails.  This is not required to
correctly deal with the complication, but is an optimisation to reduce the
parser's memory usage.  On the occasions the author has observed this
complication occurring there have been a huge number of bounce notification
mails generated: if every bounce notification mail was cached it would
dramatically increase the memory requirements of the parser.  The cache of
bounce notification mails will be pruned whenever the parser's state is
saved, though if the size of the cache ever becomes a problem it could be
pruned periodically to keep it in check.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally
mails that are in the process of being delivered will be deleted.  This
results in the log lines from the delivery agent (\daemon{local},
\daemon{virtual} or \daemon{smtp}) appearing in the log file
\textit{after\/} the mail has been removed from the state tables and saved
in the database.  The \action{DELETE} action adds deleted mails to a cache,
which is checked by the \action{SAVE\_DATA} action, and the current log
line discarded if the following conditions are met:

### 2009-03-26 doc/logparser-implementation.tex
Added a \action{} command to have consistent formatting of action names. Changes \parsernames{} to have an ' before the s.

    \item [restriction\_name] The name of the Postfix restriction that
        caused the mail to be rejected.  Only applicable to rules that
        recognise rejection log lines, i.e.\ rules that have an action of
        \texttt{DELIVERY\_REJECTED}; other rules should not set this field.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be false for a real rejection, or true if the log
        line was a warning.  This field should be ignored if the result is
        not a rejection, i.e.\ the action field of the associated rule is
        not \texttt{DELIVERY\_REJECTED}.

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up data structures that will later be used by actions,
including several state tables.  Data about the connections and mails being
processed is held in \texttt{connections} and \texttt{queueids}
respectively.  The other state tables used when working around
complications in parsing: \texttt{timeout\_queueids} is used when dealing
with connections that time out during the DATA phase (\sectionref{timeouts
during data phase}); \texttt{bounce\_queueids} is part of the solution to
bounce notification mails being delivered before their creation is logged
(\sectionref{Bounce notification mails delivered before their creation is
logged}); and \texttt{postsuper\_deleted\_queueids} caches information
about mails recently deleted by the administrator so that subsequent log
lines processed by the SAVE\_DATA action can be discarded (\sectionref{Mail
deleted before delivery is attempted}).

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt.  The next most common action is, perhaps
surprisingly, \texttt{UNINTERESTING}: this action does nothing when
invoked, allowing uninteresting log lines to be parsed without any effects
(it does not imply that the input is ungrammatical or unparsed).  Generally
rules specifying the \texttt{UNINTERESTING} action recognise log lines that
are not associated with a specific mail, e.g.\ notices about configuration
files changing; these log lines are recognised and processed so that the
framework can warn about unrecognised log lines, informing the user that
they need to add to or extend their ruleset.  Most of the remaining actions
have only one or two associated rules: some are specialised actions
required to address a deficiency in the log files, or a complication that
arises during parsing; other actions will only ever have one log line
variant, e.g.\ all log lines showing that a remote client has connected are
recognised by a single rule and handled by the \texttt{CONNECT} action.
\parsernames{} actions are described in detail in \sectionref{actions in
detail in implementation}, and \sectionref{adding new actions in
implementation} covers the process of adding a new action.

@@

This section explains the actions found in \parsername{}; it may help to
revisit the flow chart in \sectionref{flow chart} to see how a mail passes
from one action to another as its log lines are recognised.  The words
\textit{mail\/} and \textit{connection\/} are used in the actions
descriptions below because they are less unwieldy and more specific than
\textit{state table entry\/}; a connection becomes a mail during the
\texttt{CLONE} action, which processes Postfix accepting a delivery
attempt, and the data structure moves from the connections state table to
the queueids state table.

@@

    \item [DELETE] Deals with mail deleted using Postfix's administrative
        command \daemon{postsuper}.  This action adds a dummy recipient
        address if required (see \sectionref{Mail deleted before delivery
        is attempted}), then invokes the COMMIT action to save the mail to
        the database.

@@

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days, it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check,
        implemented to deal with the complication described in
        \sectionref{out of order log lines}; this action tags the mail as
        having expired, so the COMMIT action will skip the valid
        combinations check.

    \item [MAIL\_BOUNCED] This action behaves exactly like the
        \texttt{SAVE\_DATA} action; it saves all data captured by the
        recognising rule's regex, and does nothing more.  It is a separate
        action to distinguish delivery attempts that bounce from other
        delivery attempts.

    \item [MAIL\_DISCARDED] Sometimes mail is discarded, e.g.\ mail
        submitted locally that is larger than the limit configured by the
        administrator.  This action is used for those mails; it invokes the
        \texttt{COMMIT} action, but is a separate action to simplify
        further analysis.

@@

    \item [MAIL\_SENT] This action behaves exactly like the
        \texttt{SAVE\_DATA} action; it saves all data captured by the
        recognising rule's regex, and does nothing more.  It is a separate
        action to distinguish successful delivery attempts from other
        delivery attempts.

@@

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  A new mail will be
        created, although out of order log entries may have caused it to
        already exist, as documented in \sectionref{pickup logging after
        cleanup}.

@@

    \item [SAVE\_DATA] Every action that can locate the correct entry in
        the state tables saves any data captured by the recognising rule's
        regex to it.  The \texttt{SAVE\_DATA} saves data in this way but
        does not do anything else; it is invoked for log lines that contain
        useful data but do not require any further processing.

@@

    \item [TRACK] Track a mail when it is re-injected for forwarding to
        another mail server; this happens when a local address is aliased
        to a remote address (see \sectionref{tracking re-injected mail} for
        a full explanation).  \texttt{TRACK} will be called when dealing
        with the parent mail, and will create the child mail if necessary.
        \texttt{TRACK} checks if the child has already been tracked, either
        by this parent or by another parent, and issues appropriate
        warnings in either case.

@@

    \textbf{Field}      & \textbf{Value}                                    \\
    name                & Unknown sender domain                             \\
    description         & We do not accept mail from unknown domains        \\
    restriction\_name   & reject\_unknown\_sender\_domain                   \\
    program             & \daemon{smtpd}                                    \\
    regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                        & \verb!Sender address rejected: Domain not found;! \\
                        & \verb!from=<\k<sender>> to=<(__RECIPIENT__)> !    \\
                        & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
    result\_data        &                                                   \\
    connection\_data    &                                                   \\
    action              & DELIVERY\_REJECTED                                \\
    hits                & 0                                                 \\
    hits\_total         & 0                                                 \\
    priority            & 0                                                 \\
    %cluster\_group      & 400                                               \\

@@

There are other attributes that may be required to make a rule work
correctly: connection\_data, result\_data, priority, or restriction\_name.
In general it will only become clear that connection\_data or result\_data
are required when \parsername{} warns about an entry in the connections or
results tables that is missing some required fields, because values for the
missing fields are not present in any of the log lines.  E.g.\ the rule
recognising the \daemon{pickup} component processing a mail sets the
client\_hostname to \texttt{localhost} and the client\_ip to
\texttt{127.0.0.1}, because the mail originates on the local server.  If
the new rule deliberately uses the architecture's overlapping rules feature
it will be obvious that the priority field needs to be set, on this rule
and possibly others; the priority field may be needed on unintentionally
overlapping rules too, but that is more difficult to determine.  Finally
the restriction\_name field should be set when the rule's action is
\texttt{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
clear from the content of the log line.

@@

    \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
        dies or is killed; see the SMTPD\_DIED action in
        \sectionref{actions in detail in implementation}.

>>>>

    \item [restriction\_name] The name of the Postfix restriction that
        caused the mail to be rejected.  Only applicable to rules that
        recognise rejection log lines, i.e.\ rules that have an action of
        \action{DELIVERY\_REJECTED}; other rules should not set this field.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be false for a real rejection, or true if the log
        line was a warning.  This field should be ignored if the result is
        not a rejection, i.e.\ the action field of the associated rule is
        not \action{DELIVERY\_REJECTED}.

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up data structures that will later be used by actions,
including several state tables.  Data about the connections and mails being
processed is held in \texttt{connections} and \texttt{queueids}
respectively.  The other state tables used when working around
complications in parsing: \texttt{timeout\_queueids} is used when dealing
with connections that time out during the DATA phase (\sectionref{timeouts
during data phase}); \texttt{bounce\_queueids} is part of the solution to
bounce notification mails being delivered before their creation is logged
(\sectionref{Bounce notification mails delivered before their creation is
logged}); and \texttt{postsuper\_deleted\_queueids} caches information
about mails recently deleted by the administrator so that subsequent log
lines processed by the \action{SAVE\_DATA} action can be discarded
(\sectionref{Mail deleted before delivery is attempted}).

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \action{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt.  The next most common action is, perhaps
surprisingly, \action{UNINTERESTING}: this action does nothing when
invoked, allowing uninteresting log lines to be parsed without any effects
(it does not imply that the input is ungrammatical or unparsed).  Generally
rules specifying the \action{UNINTERESTING} action recognise log lines that
are not associated with a specific mail, e.g.\ notices about configuration
files changing; these log lines are recognised and processed so that the
framework can warn about unrecognised log lines, informing the user that
they need to add to or extend their ruleset.  Most of the remaining actions
have only one or two associated rules: some are specialised actions
required to address a deficiency in the log files, or a complication that
arises during parsing; other actions will only ever have one log line
variant, e.g.\ all log lines showing that a remote client has connected are
recognised by a single rule and handled by the \action{CONNECT} action.
\parsernames{} actions are described in detail in \sectionref{actions in
detail in implementation}, and \sectionref{adding new actions in
implementation} covers the process of adding a new action.

@@

This section explains the actions found in \parsername{}; it may help to
revisit the flow chart in \sectionref{flow chart} to see how a mail passes
from one action to another as its log lines are recognised.  The words
\textit{mail\/} and \textit{connection\/} are used in the actions
descriptions below because they are less unwieldy and more specific than
\textit{state table entry\/}; a connection becomes a mail during the
\action{CLONE} action, which processes Postfix accepting a delivery
attempt, and the data structure moves from the connections state table to
the queueids state table.

@@

    \item [DELETE] Deals with mail deleted using Postfix's administrative
        command \daemon{postsuper}.  This action adds a dummy recipient
        address if required (see \sectionref{Mail deleted before delivery
        is attempted}), then invokes the \action{COMMIT} action to save the
        mail to the database.

@@

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days, it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check,
        implemented to deal with the complication described in
        \sectionref{out of order log lines}; this action tags the mail as
        having expired, so the \action{COMMIT} action will skip the valid
        combinations check.

    \item [MAIL\_BOUNCED] This action behaves exactly like the
        \action{SAVE\_DATA} action; it saves all data captured by the
        recognising rule's regex, and does nothing more.  It is a separate
        action to distinguish delivery attempts that bounce from other
        delivery attempts.

    \item [MAIL\_DISCARDED] Sometimes mail is discarded, e.g.\ mail
        submitted locally that is larger than the limit configured by the
        administrator.  This action is used for those mails; it invokes the
        \action{COMMIT} action, but is a separate action to simplify
        further analysis.

@@

    \item [MAIL\_SENT] This action behaves exactly like the
        \action{SAVE\_DATA} action; it saves all data captured by the
        recognising rule's regex, and does nothing more.  It is a separate
        action to distinguish successful delivery attempts from other
        delivery attempts.

@@

    \item [PICKUP] The \action{PICKUP} action corresponds to the
        \daemon{pickup} service dealing with a locally submitted mail.  A
        new mail will be created, although out of order log entries may
        have caused it to already exist, as documented in
        \sectionref{pickup logging after cleanup}.

@@

    \item [SAVE\_DATA] Every action that can locate the correct entry in
        the state tables saves any data captured by the recognising rule's
        regex to it.  The \action{SAVE\_DATA} saves data in this way but
        does not do anything else; it is invoked for log lines that contain
        useful data but do not require any further processing.

@@

    \item [TRACK] Track a mail when it is re-injected for forwarding to
        another mail server; this happens when a local address is aliased
        to a remote address (see \sectionref{tracking re-injected mail} for
        a full explanation).  \action{TRACK} will be called when dealing
        with the parent mail, and will create the child mail if necessary.
        \action{TRACK} checks if the child has already been tracked, either
        by this parent or by another parent, and issues appropriate
        warnings in either case.

@@

    \textbf{Field}      & \textbf{Value}                                    \\
    name                & Unknown sender domain                             \\
    description         & We do not accept mail from unknown domains        \\
    restriction\_name   & reject\_unknown\_sender\_domain                   \\
    program             & \daemon{smtpd}                                    \\
    regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                        & \verb!Sender address rejected: Domain not found;! \\
                        & \verb!from=<\k<sender>> to=<(__RECIPIENT__)> !    \\
                        & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
    result\_data        &                                                   \\
    connection\_data    &                                                   \\
    action              & \action{DELIVERY\_REJECTED}                       \\
    hits                & 0                                                 \\
    hits\_total         & 0                                                 \\
    priority            & 0                                                 \\
    %cluster\_group      & 400                                               \\

@@

There are other attributes that may be required to make a rule work
correctly: connection\_data, result\_data, priority, or restriction\_name.
In general it will only become clear that connection\_data or result\_data
are required when \parsername{} warns about an entry in the connections or
results tables that is missing some required fields, because values for the
missing fields are not present in any of the log lines.  E.g.\ the rule
recognising the \daemon{pickup} component processing a mail sets the
client\_hostname to \texttt{localhost} and the client\_ip to
\texttt{127.0.0.1}, because the mail originates on the local server.  If
the new rule deliberately uses the architecture's overlapping rules feature
it will be obvious that the priority field needs to be set, on this rule
and possibly others; the priority field may be needed on unintentionally
overlapping rules too, but that is more difficult to determine.  Finally
the restriction\_name field should be set when the rule's action is
\action{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
clear from the content of the log line.

@@

    \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
        dies or is killed; see the \action{SMTPD\_DIED} action in
        \sectionref{actions in detail in implementation}.
